Using VariantSpark to learn population structure from genomic profiles

### 2015-10-26 doc/publications/Recomb15/recomb15.tex
chr22 results

\subsubsection{Random Forest does not scale to whole genome analysis and is outperformed by logistic regression}
In this section we compare RF against \kMeans\ and the LR.  
\marginpar{TODO elaborate on the error} 
Despite using {\sc Spark} and {\sc MLlib}, RF is not able to scale to the full dataset of 1000 individuals and 40
Million variables, due to a limitation of the Java virtual machine (maximum array size) that causes {\sc Mllib}'s
random forest implementation to fail. This error is due to the number of features present in each sample,
with a potential solution being to limit the features through feature selection, or simply using a subset of the dataset.
We therefore use only chromosome 22 (494,328 alleles) for the comparison in this section.  We train a RF
and LR on this subset of genomic profiles from the 1000 Genomes data using the annotated labels of all
super-populations.  We perform 10-fold cross-validation and report the mean accuracy over all training and testing folds
along with the standard error.  We also perform 10 runs of \kMeans{} with 4 clusters (\(k=4\)) and report the mean accuracy of the
resulting models along with the standard error.

\marginpar{TODO elaborate on accuracy measures}
As reported in our earlier work, \kMeans\ is not able to fit the dataset accurately.
Although capable of fitting a relatively accurate model (ARI=0.832), the average ARI over ten models is  \(0.553 \pm 0.07\), see table~\ref{RF}.
RF achieves a better fitting model (XX\%, ARI=Y) but is outperformed by the simple logistic regression, with an ARI of \(0.923 \pm 0.02\).
This indicates that a multivariate approach may not be needed to accurately fit ethnicity.

\begin{table}
\caption{Cross-validated results, mean accuracy (and standard error) on variants from chromosome 22 of 4 super-populations.}
\begin{center}
\renewcommand{\arraystretch}{1.4}
\setlength\tabcolsep{3pt}
\begin{tabular}{lcccc}
\hline\noalign{\smallskip}
method  & test folds & training folds & hold out \\
RF  & X (Y) & X (Y) & X (Y) \\
\kMeans & X (Y) & X (Y) & X (Y) \\
LR & X (Y) & X (Y) & X (Y) \\
\noalign{\smallskip}
\hline
\label{RF}
\end{tabular}
\end{center}
\end{table}

@@

\begin{table}
\caption{Cross-validated results, mean accuracy (and standard error) on 3 populations.}
\begin{center}
\renewcommand{\arraystretch}{1.4}
\setlength\tabcolsep{3pt}
\begin{tabular}{lcccc}
\hline\noalign{\smallskip}
method  & test folds & training folds & hold out \\
RF  & X (Y) & X (Y) & X (Y) \\
\kMeans & X (Y) & X (Y) & X (Y) \\
deep learning & X (Y) & X (Y) & X (Y) \\
\noalign{\smallskip}
\hline
\end{tabular}
\end{center}
\end{table}

>>>>

\subsubsection{Random Forest does not scale to whole genome analysis and is outperformed by logistic regression}
In this section we compare RF against \kMeans\ and the LR.  
\marginpar{TODO elaborate on the error} 
Despite using {\sc Spark} and {\sc MLlib}, RF is not able to scale to the full dataset of 1000 individuals and 40
Million variables, due to a limitation of the Java virtual machine (maximum array size) that causes {\sc Mllib}'s
random forest implementation to fail. This error is due to the number of features present in each sample,
with a potential solution being to limit the features through feature selection, or simply using a subset of the dataset.
We therefore use only chromosome 22 (494,328 alleles) for the comparison in this section.  We train a RF
and LR on this subset of genomic profiles from the 1000 Genomes data using the annotated labels of all
super-populations.  We perform 10-fold cross-validation and report the mean error over all training and testing folds
along with the standard error of the mean (SEM). We also use the Adjusted Rand Index as a metric, as it allows us to
compare the classification results with the \kMeans{} clustering result. We also perform 10 runs of \kMeans{} with 4 clusters (\(k=4\)) and report the mean accuracy of the
resulting models along with the standard error.

\marginpar{TODO elaborate on accuracy measures}
As reported in our earlier work, \kMeans\ is not able to fit the dataset accurately.
Although capable of fitting a relatively accurate model \(ARI=0.837\), the average ARI over ten models is  \(0.596 \pm 0.067\), see table~\ref{RF}.
RF achieves a better fitting model \(ARI = 0.757 \pm 0.011\) but is outperformed by the simple logistic regression, with an ARI of \(0.923 \pm 0.015\).
This indicates that a multivariate approach may not be needed to accurately fit ethnicity.

\begin{table}
\caption{Cross-validated results, mean accuracy (and standard error) on variants from chromosome 22 of 4 super-populations.}
\begin{center}
\renewcommand{\arraystretch}{1.4}
\setlength\tabcolsep{3pt}
\begin{tabular}{lcccc}
\hline\noalign{\smallskip}
method  & ARI & error (\%) & hold out \\
RF  & 0.757 (0.011) &0.13 (0.007) & X (Y) \\
\kMeans & 0.596 (0.066) & -- & X (Y) \\
LR & 0.923 (0.015) & 0.034 (0.006) & X (Y) \\
\noalign{\smallskip}
\hline
\label{RF}
\end{tabular}
\end{center}
\end{table}

@@

\begin{table}
\caption{Cross-validated results, mean accuracy (and standard error) on 3 populations.}
\begin{center}
\renewcommand{\arraystretch}{1.4}
\setlength\tabcolsep{3pt}
\begin{tabular}{lcccc}
\hline\noalign{\smallskip}
method  & ARI & training folds & hold out \\
RF  & X (Y) & X (Y) & X (Y) \\
\kMeans & X (Y) & X (Y) & X (Y) \\
deep learning & 0.976 & X (Y) & X (Y) \\
\noalign{\smallskip}
\hline
\end{tabular}
\end{center}
\end{table}
